### 2015-09-12 paper/main.tex
Fix #7 - discuss firms vs establishments

Our findings support the main model predictions, with tariff barriers stifling domestic competition, leading to higher producer prices and markups as well as lower productivity. In the immediate years after the free-trade agreement when tariff barriers are reduced, relative prices and markups decrease as relative productivity increases, thus giving rise to competitive effects. The results in the long-run, however, are not as clear cut, with some effects reversing as predicted by the model while some effects persist. This is also confirmed by directly looking at the reaction of industries with different entry barriers to changes in trade openness.  
\vspace{0.5cm}

@@

\subsection{Further Changes (work in progress!)}
THINK CLOSELY ABOUT THIS PARAGRAPH! \\
Following the arguments in \citet{Trefler2004}, we want to exploit the nature of NAFTA being close to a natural experiment and hence try to identify the effect of the policy measures (i.e. the changes in tariff rates) separately from the effects of trade openness in general. There is evidence that trade openness -- measured by the import penetration of a certain country or industry, as in \citet{Chen2009} -- is affected by a number of external forces, including oil prices \citep{Kilian2009}, credit conditions \citep{Chor2012}, shared culture and language between countries and many more. Therefore, we deconstruct the iceberg costs of trade into two parts: $\tau^{lh}=\frac{Tr^{lh}}{\theta^{lh}}$, where $Tr^{lh}>1$ is the tariff rate for trade between countries $l$ and $h$ and $\theta^{lh}>1$ captures the additional costs of trade imposed by the aforementioned factors. Then, the import penetration can be viewed as a measure of $\theta^{lh}$, as in Chen et al. (2009), while a carefully constructed tariff measure can directly account for the effect of NAFTA-mandated changes in the trade environment. Obviously, the import penetration in a sector depends on the tariff measure as well, so an instrumental variable approach has to be taken in order to identify this effect. We defer the discussion of the construction of the tariff measure and the choice of instruments to the following chapter. Prior work of \citet{Bernard2006} suggests that tariff rates throughout the 1980s, at an average level between four and five percent, accounted for about the same fraction of trade costs as costs directly attached to shipping the good, i.e. freight and insurance, so we expect them to have a sizable impact on trade flows between countries and hence the competitive environment. \vspace{0.5cm} \\

@@

%Furthermore, we rely on work by \citet{Loecker2012}, who use a production function with at least one variable input to identify markups from the output elasticity with respect to the variable input and the expenditure share of total revenue of a firm. We also discuss the specifics of this procedure and the data needed to implement it in the following chapter.

\section{Application}\label{sec:app}
Starting from the equations for prices, productivity and markups derived within the Melitz-Ottaviano framework, Chen et al. then proceed by constructing log-linearized equations to estimate. The estimation equation for prices is given by:
\begin{align}
\begin{split}\label{eq:chen-estimation-prices}
\Delta \ln \left( \frac{\bar{p}_{it}}{\bar{p}_{it}^*} \right) = &\beta_0 + \beta_1 \Delta \ln \theta_{it} + \beta_2 \Delta \ln \theta_{it}^* + \beta_3 \Delta \ln D_{it} + \beta_4 \Delta \ln D_{it}^* \\ &+ \gamma \left[ \ln \left( \frac{\bar{p}_{it-1}}{\bar{p}_{it-1}^*} \right) + \delta_0 + \delta_1 \ln L_{t-1} + \delta_2\ln L_{t-1}^* + \delta_3 \ln  \theta_{i,t-1} + \delta_4 \ln  \theta_{i,t-1}^* \right] + \varepsilon_{ijt} 
\end{split}\end{align}
In the above equation, the number of firms serving the domestic market, $N$, has been replaced by the more readily observable number of domestic firms producing for the domestic market, $D$, where $D=N \left( \frac{c_D}{c_M} \right)^k$. The short--run dynamics are estimated in the first part of the equation, with regressors expressed in first differences. The long run is represented by the term in brackets. From the perspective of this model, we would expect $\beta_1<0$, an increase in domestic openness lowers relative prices in the short-run, and hence $\beta_2>0$ as well as $\beta_3<0$ and $\beta_4>0$ -- a higher number of domestic producers lowers the domestic price level while the number of foreign producers has the opposite effect.

The model specification used in this analysis is quite similar to that of \citet{Chen2009}, however, we have included an arguably better measurement of domestic openness: tariffs imposed on foreign products. The estimation equation for prices utilized in this analysis is given by: 
\begin{align}
\begin{split}\label{eq:gw-estimation-prices}
\Delta \ln \left( \frac{\bar{p}_{it}}{\bar{p}_{it}^*} \right) = &\beta_0 + \beta_1 \Delta \tau_{it} + \beta_2 \Delta \tau_{it}^* + \beta_4 \Delta \ln D_{it} + \beta_5 \Delta \ln D_{it}^* \\ 
&+ \gamma \left[ \ln \left( \frac{\bar{p}_{it-1}}{\bar{p}_{it-1}^*} \right) + \delta_0 + \delta_1 \ln L_{t-1} + \delta_2\ln L_{t-1}^* + \delta_3  \tau_{i,t-1} + \delta_4  \tau_{i,t-1}^* \right] + \varepsilon_{ijt} 
\end{split}
\end{align}
Similarly, the estimation equation that is utilized in this analysis for markups is given by:
\begin{align}
\begin{split}\label{eq:gw-estimation-markup}
\Delta \ln \left( \frac{\bar{\mu}_{it}}{\bar{\mu}_{it}^*} \right) = &\beta_0 + \beta_1 \Delta \tau_{it} + \beta_2 \Delta \tau_{it}^* + \beta_3 \Delta \ln \theta_{it} + \beta_4 \Delta \ln \theta_{it}^* + \beta_5 \Delta \ln D_{it} + \beta_6 \Delta \ln D_{it}^*\\ 
&+ \gamma \left[ \ln \left( \frac{\bar{\mu}_{it-1}}{\bar{\mu}_{it-1}^*} \right) + \delta_0 + \delta_1 \ln L_{t-1} + \delta_2\ln L_{t-1}^* + \delta_3  \tau_{i,t-1} + \delta_4  \tau_{i,t-1}^* \right. \\
&+ \delta_5 \ln  \theta_{i,t-1} +  \left. \delta_6 \ln  \theta_{i,t-1}^* \right] + \varepsilon_{ijt} 
\end{split}
\end{align}
The effect of tariffs, openness, number of firms and market size on productivity is estimated by:
\begin{align}
\begin{split}\label{eq:gw-estimation-productivity}
\Delta \ln \left( \frac{\bar{z}_{it}}{\bar{z}_{it}^*} \right) =&\beta_0 + \beta_1 \Delta \tau_{it} + \beta_2 \Delta \tau_{it}^* + \beta_3 \Delta \ln \theta_{it} + \beta_4 \Delta \ln \theta_{it}^* + \beta_5 \Delta \ln D_{it} + \beta_6 \Delta \ln D_{it}^*\\
&+ \gamma \left[ \ln \left( \frac{z_{it-1}}{z_{it-1}^*} \right) + \delta_0 + \delta_1 \ln L_{t-1} + \delta_2\ln L_{t-1}^* + \delta_3  \tau_{i,t-1} + \delta_4  \tau_{i,t-1}^*  \right. \\ 
&+ \delta_5 \ln  \theta_{i,t-1}+ \left. \delta_6 \ln  \theta_{i,t-1}^* + \delta_7 \ln w_{i,t-1} +  \delta_8 \ln w_{i,t-1}^* \right] + \varepsilon_{ijt} 
\end{split}
\end{align}
where $\delta_7$ and $\delta_8$ capture the effects of changes in nominal wages in the long run. The intercepts $\beta_0$ are introduced to capture differences in country--specific technology as Chen et al. depart from the baseline Melitz-Ottaviano model by allowing for such differences.

@@

The database we utilize covers the period 1994-2006 for NAFTA member countries -- Canada, Mexico, and the US -- and nine manufacturing sectors. For our factory gate price data, we use the producer price index (PPI) as reported by CANSIM, the Banco de Mexico, and the U.S. Bureau of Labor Statistics, respectively. All indices are normalized to equal 100 in 2003. While the majority of the manufacturing data collected is reported at the two-digit level according to ISIC Rev. 3, we aggregate all data according to ISIC Rev. 2 in order to keep consistency throughout the analysis (see Table \ref{tb:isic_class} for the manufacturing classification breakdown).

@@

As discussed in the previous section, for all of the log-linearized equations, we replace the number of firms serving the domestic market, $N$, with the number of domestic firms producing for the domestic market, $D$. Unfortunately, this data is not available for all three countries during the specified time period, and thus we utilize the number of establishments, which will always be higher than the firm count as each firm may have multiple establishments; for this reason, the coefficient on this variable will always be at the upper bound and must be evaluated with caution. The domestic establishment data is made available by CANSIM, the U.S. Bureau of Labor Statistics, and the Instituto Nacional de Estadistica.

@@

\begin{center}
\begin{table*}[ht] \caption{Comparative Statics -- Model Predictions}\label{tb:comp_stats}
\vspace{0.5cm}
{\small
\hfill{}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
\textbf{Regressor}& \multicolumn{6}{|c|}{\textbf{Dependent Variables}}\\
\cline{2-7}
\hline
& $\bar{p}_{sr}$ & $\bar{p}_{lr}$ & $\mu_{sr}$ & $\mu_{lr}$ & $z_{sr}$ & $z_{lr}$ \\ \hline
$\tau$     & +  & -- & +  & -- & -- & + \\ \hline
$\tau^*$   & -- & +  & -- & +  & +  & --\\ \hline
$\theta$   & -- & +  & -- & +  & +  & --\\ \hline
$\theta^*$ & +  & -- & +  & -- & -- & + \\ \hline
$D$        & -- &    & -- &    & +  &   \\ \hline
$D^*$      & +  &    & +  &    & -- &   \\ \hline
$L$        &    & -- &    & -- &    & + \\ \hline
$L^*$      &    & +  &    & +  &    & --\\
\hline
\end{tabular}}
\hfill{}
\end{table*}
\end{center}

>>>>

Our findings support the main model predictions, with tariff barriers stifling domestic competition, leading to higher producer prices and markups as well as lower productivity. In the immediate years after the free-trade agreement when tariff barriers are reduced, relative prices and markups decrease as relative productivity increases, thus giving rise to competitive effects. The results in the long-run, however, are not as clear cut, with some effects reversing as predicted by the model while some effects persist. This is also confirmed by directly looking at the reaction of industries with different entry barriers to changes in trade openness. \\
\vspace{0.5cm} \\

@@

\subsection{The Role of Market Entry}
Following the arguments in \citet{Trefler2004}, we want to exploit the nature of NAFTA being close to a natural experiment and hence try to identify the effect of the policy measures (i.e. the changes in tariff rates) separately from the effects of trade openness in general. There is evidence that trade openness -- measured by the import penetration of a certain country or industry, as in \citet{Chen2009} -- is affected by a number of external forces, including oil prices \citep{Kilian2009}, credit conditions \citep{Chor2012}, shared culture and language between countries and many more. Therefore, we deconstruct the iceberg costs of trade into two parts: $\tau^{lh}=\frac{Tr^{lh}}{\theta^{lh}}$, where $Tr^{lh}>1$ is the tariff rate for trade between countries $l$ and $h$ and $\theta^{lh}>1$ captures the additional costs of trade imposed by the aforementioned factors. Then, the import penetration can be viewed as a measure of $\theta^{lh}$, as in Chen et al. (2009), while a carefully constructed tariff measure can directly account for the effect of NAFTA-mandated changes in the trade environment. Obviously, the import penetration in a sector depends on the tariff measure as well, so an instrumental variable approach has to be taken in order to identify this effect. We defer the discussion of the construction of the tariff measure and the choice of instruments to the following chapter. Prior work of \citet{Bernard2006} suggests that tariff rates throughout the 1980s, at an average level between four and five percent, accounted for about the same fraction of trade costs as costs directly attached to shipping the good, i.e. freight and insurance, so we expect them to have a sizable impact on trade flows between countries and hence the competitive environment. \vspace{0.5cm} \\

@@

\section{Application}\label{sec:app}
Starting from the equations for prices, productivity and markups derived within the Melitz-Ottaviano framework, we can derive estimable log-linearized equations analogous to those in Chen et al. The estimation equation for prices is given by:
\begin{align}
\begin{split}\label{eq:gw-estimation-prices}
\Delta \ln \left( \frac{\bar{p}_{it}}{\bar{p}_{it}^*} \right) = &\beta_0 + \beta_1 \Delta \ln \tau_{it} + \beta_2 \Delta \ln \tau_{it}^* + \beta_3 \Delta \ln D_{it} + \beta_4 \Delta \ln D_{it}^* \\ &+ \gamma \left[ \ln \left( \frac{\bar{p}_{it-1}}{\bar{p}_{it-1}^*} \right) + \delta_0 + \delta_1 \ln L_{t-1} + \delta_2\ln L_{t-1}^* + \delta_3 \ln  \tau_{i,t-1} + \delta_4 \ln  \tau_{i,t-1}^* \right] + \varepsilon_{ijt} 
\end{split}\end{align}
In the above equation, the number of firms serving the domestic market, $N$, has been replaced by the more readily observable number of domestic firms producing for the domestic market, $D$, where $D=N \left( \frac{c_D}{c_M} \right)^k$. The short--run dynamics are estimated in the first part of the equation, with regressors expressed in first differences. The long run is represented by the term in brackets. From the perspective of this model, we would expect $\beta_1>0$, an increase in domestic import tariffs increases relative prices in the short-run, and correspondingly $\beta_2>0$. The model predicts a dampening effect of the number of domestic firms on domestic prices, which should be reflected by $\beta_3<0$, and the opposite for foreign firms, $\beta_4<0$. As we expect the coefficient on the error correction term, $\gamma$, to be negative, a reversal the

As previously discussed, all aggregate variables (prices, markups, productivity) are ultimately functions of the cost-cutoff level $c_D$, leading to very similar estimation equations for our three dependent variables. The equation markups is:
\begin{align}
\begin{split}\label{eq:gw-estimation-markup}
\Delta \ln \left( \frac{\bar{\mu}_{it}}{\bar{\mu}_{it}^*} \right) = &\beta_0 + \beta_1 \Delta \tau_{it} + \beta_2 \Delta \tau_{it}^* + \beta_3 \Delta \ln \theta_{it} + \beta_4 \Delta \ln \theta_{it}^* + \beta_5 \Delta \ln D_{it} + \beta_6 \Delta \ln D_{it}^*\\ 
&+ \gamma \left[ \ln \left( \frac{\bar{\mu}_{it-1}}{\bar{\mu}_{it-1}^*} \right) + \delta_0 + \delta_1 \ln L_{t-1} + \delta_2\ln L_{t-1}^* + \delta_3  \tau_{i,t-1} + \delta_4  \tau_{i,t-1}^* \right] + \varepsilon_{ijt} 
\end{split}
\end{align}
The effect of tariffs, openness, number of firms and market size on productivity is estimated by:
\begin{align}
\begin{split}\label{eq:gw-estimation-productivity}
\Delta \ln \left( \frac{\bar{z}_{it}}{\bar{z}_{it}^*} \right) =&\beta_0 + \beta_1 \Delta \tau_{it} + \beta_2 \Delta \tau_{it}^*  + \beta_3 \Delta \ln D_{it} + \beta_4 \Delta \ln D_{it}^*\\
&+ \gamma \left[ \ln \left( \frac{z_{it-1}}{z_{it-1}^*} \right) + \delta_0 + \delta_1 \ln L_{t-1} + \delta_2\ln L_{t-1}^* + \delta_3  \tau_{i,t-1} + \delta_4  \tau_{i,t-1}^*  \right. \\ 
&+ \left. \delta_5 \ln w_{i,t-1} +  \delta_6 \ln w_{i,t-1}^* \bigg] \right. + \varepsilon_{ijt} 
\end{split}
\end{align}
where $\delta_7$ and $\delta_8$ capture the effects of changes in nominal wages in the long run. The intercepts $\beta_0$ are introduced to capture differences in country--specific technology as Chen et al. depart from the baseline Melitz-Ottaviano model by allowing for such differences. While in the baseline, these vary by country-pair, we check the robustness of the specification by allowing fixed effects at a sectoral level.

@@

The database we utilize covers the period 1994-2006 for NAFTA member countries -- Canada, Mexico, and the US -- and nine manufacturing sectors. For our factory gate price data, we use the producer price index (PPI) as reported by CANSIM, the Banco de Mexico, and the U.S. Bureau of Labor Statistics, respectively. All indices are normalized to equal 100 in 2003. While the majority of the manufacturing data collected is reported at the two-digit level according to ISIC Rev. 3, we aggregate all data according to NACIS in order to keep consistency throughout the analysis (see Table \ref{tb:isic_class} for the manufacturing classification breakdown). While the main part of our analysis rests on data aggregated at the two-digit level, we were also able to assemble a data set at the four-digit level, which we will use to check the robustness of our results.

@@

As discussed in the previous section, for all of the log-linearized equations, we replace the number of firms serving the domestic market, $N$, with the number of domestic firms producing for the domestic market, $D$. Unfortunately, this data is not available for all three countries during the specified time period, and thus we utilize the number of establishments, which will always be higher than the firm count as each firm may have multiple establishments. As long as the average number of establishments per firm remains constant, this should not present a problem, as our model is estimated in first differences. It is however not obvious that this relationship will remain constant in response to a trade liberalization. In fact, the main channel through which welfare gains arise in the model is the reallocation of production from unproductive to more productive firms, with less productive firms exiting the market and more productive firms expanding. If this displacement happens through larger firms taking over establishments of less productive ones, we would expect the number of establishments to stay constant, while the number of domestic producers falls (i.e. the number of establishments per firm increases). If on the other hand larger firms are simply able to expand production in existing establishments, this effect would be absent.

@@

\begin{center}
\begin{table*}[ht] \caption{Comparative Statics -- Model Predictions}\label{tb:comp_stats}
\vspace{0.5cm}
{\small
\hfill{}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
\textbf{Regressor}& \multicolumn{6}{|c|}{\textbf{Dependent Variables}}\\
\cline{2-7}
\hline
& $\bar{p}_{sr}$ & $\bar{p}_{lr}$ & $\mu_{sr}$ & $\mu_{lr}$ & $z_{sr}$ & $z_{lr}$ \\ \hline
$\tau$     & +  & -- & +  & -- & -- & + \\ \hline
$\tau^*$   & -- & +  & -- & +  & +  & --\\ \hline
$D$        & -- &    & -- &    & +  &   \\ \hline
$D^*$      & +  &    & +  &    & -- &   \\ \hline
$L$        &    & -- &    & -- &    & + \\ \hline
$L^*$      &    & +  &    & +  &    & --\\
\hline
\end{tabular}}
\hfill{}
\end{table*}
\end{center}
